---
title: "Status of Benzene Emissions and Monitoring in the UK"
author: "Lucy Webster"
format:
  html:
    toc: true
    toc-location: left
    crossrefs-hover: false
  docx:
    toc: true
  pdf:
    toc: true
    fig-align: center
    fig-width: 6.5
    fig-height: 3.8
    geometry:
      - top=2.5cm
      - bottom=2.5cm
      - left=2.5cm
      - right=2.5cm
editor: source
---

```{r setup}
#| echo: false
#| message: false
#| warning: false


# Loading in r packages required and setting the quarto doc to look at the project root 




knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(
    out.width = "100%", # enough room to breath
    fig.width = 6,     # reasonable size
    fig.asp = 0.718,   # golden ratio
    fig.align = "center" # mostly what I want
)



library(tidyverse)
library(plotly)
library(scales)
library(sf)
library(leaflet)




base_size <- if (knitr::is_html_output()) 14 else 10          # smaller text on A4

knitr::opts_chunk$set(
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png",   # fixes µ and ⁻³ (see 4)
  fig.align = "center"
)

show_plot <- function(p, height = NULL) {
  if (knitr::is_html_output()) {
    ggplotly(p, tooltip = "text", height = height) |>
      config(displayModeBar = FALSE)
  } else {
    p
  }
}



```

```{r laoding in datasets}
#| echo: false
#| message: false
#| warning: false



yearly_emissions <- read_csv("data/basic_benzene_data_for_ally/total_benzene_emissions.csv")


naei_totals_raw <- read_csv(
  "data/naei_inventory_total/naei_uk_emissions_data_benzene.csv",
  na = c("", "NA", "-")
)

source_lookup <- read_csv("data/naei_inventory_total/benzene_source_lookup.csv")

# ---- Process NAEI data (moved up so every figure uses the same totals) ----
naei_totals <- naei_totals_raw |>
  filter(!str_starts(`NFR/CRT Group`, "z_")) |>          # drop memo items
  pivot_longer(`1990`:`2023`,
               names_to = "year", values_to = "benzene_emission",
               names_transform = as.numeric) |>
  left_join(source_lookup,
            by = c("NFR/CRT Group" = "nfr_code", "Source" = "source"))

# Check: stops with an error if a new NAEI release adds a source you haven't mapped
stopifnot(!anyNA(naei_totals$final_category))

# National total excluding memo items (replaces total_benzene_emissions.csv,
# which appears to include them: 62.1 kt vs 60.9 kt in 1990)
yearly_emissions <- naei_totals |>
  summarise(total_emissions_kt = sum(benzene_emission, na.rm = TRUE), .by = year)
```

## Total Emissions

The [UK National Atmospheric Emissions Inventory (NAEI)](https://naei.energysecurity.gov.uk/) produce annual estimates of the emissions of Benzene in the UK. Total emissions have fallen from 60.9 kt in 1990 to 8.0 kt in 2023, a reduction of 87%, see @fig-total-emissions. The majority of this reduction happened in the 1990s, as catalytic converters and emission controls applied to the car fleet. The single biggest year-on-year drop was between 1999 and 2000, when emissions fell from 37.8 kt to 22.1 kt. This coincided with the EU Fuel Quality Directive (98/70/EC) reducing the maximum allowable benzene content of petrol from 5% to 1% by volume from January 2000. Since around 2010 annual emissions of benzene have plateaued, and as of 2023 annual emissions were 8.0 kt.

*Emissions reported as memo items (including international maritime navigation and international aviation cruise, "z\_") were excluded, consistent with national totals reported under the UNECE Convention on Long-range Transboundary Air Pollution (CLRTAP).*

```{r}
#| label: fig-total-emissions
#| fig-cap: "Annual UK benzene emissions, 1990–2023. Excludes memo items (international shipping and international aviation cruise). Source: NAEI."
#| echo: false
#| warning: false
#| fig-height: 5

p_total <- yearly_emissions |>
    ggplot(aes(
      x = year, 
      y = total_emissions_kt
    )) +
    geom_point(aes(text = paste0("Year: ", year, "<br>",
                        "Emissions: ", round(total_emissions_kt, 1), " kt")),
               size = 0.8) +
    geom_line() +
    scale_y_continuous(
      name = "Benzene Emissions (kilotonnes)",
      expand = c(0, 0),
      breaks = seq(0, 60, 10),
      limits = c(0, 65)
    ) +
    scale_x_continuous(
      name = "Year",
      expand = c(0, 0),
      limits = c(1989.5, 2025.5),
      breaks = seq(1990, 2026, 5),
      minor_breaks = seq(1990, 2025, 1)
    )  +
    theme_minimal(base_size) +
    theme(
      legend.position = "top",
      axis.line = element_line(),
      axis.ticks = element_line()
    )


show_plot(p_total)

```

### By Emission Source

The NAEI reports benzene emissions from about 342 individual sources. For this report they have been grouped into seven broad categories (see `benzene_source_lookup.csv` for the full mapping). The figure below shows each category's share of the national total in each year.

In 1990, road transport dominated UK benzene emissions and accounted for three-quarters (75%) of the total. Its share has fallen steadily as regulation has been introduced, and by 2023 road transport contributed 12% of the total emissions (0.9 kt, down from 46 kt). No other source have seen similar levels of reduction so the mix of sources now looks different. In 2015, Domestic combustion overtook road transport as the largest single source in 2015 and made up a third (33%) of emissions in 2023. Industry (16%), shipping and other transport (13%) and waste (11%) are now all comparable to or larger than road transport.

```{r bar-chart}
#| label: fig-source-shares
#| fig-cap: "Share of UK benzene emissions by source category, 1990–2023. Sources have been grouped into seven categories (see benzene_source_lookup.csv). Source: NAEI."
#| echo: false
#| fig-height: 5

# Order categories by 2023 size so plot legends read biggest to smallest
cat_order <- naei_totals |>
  filter(year == 2023) |>
  summarise(total = sum(benzene_emission, na.rm = TRUE), .by = final_category) |>
  arrange(desc(total)) |>
  pull(final_category)

naei_totals <- naei_totals |>
  mutate(final_category = factor(final_category, levels = cat_order))


# ---- Fixed colours: one per category, matched by name ----
source_colours <- c(
  "Domestic"                   = "#FFCD12FF",  
  "Industry"                   = "#A40000FF",  
  "Shipping & other transport" = "#007E2FFF",  
  "Road transport"             = "#16317DFF",  
  "Waste"                      = "#721B3EFF", 
  "Oil, gas & fuel supply"     = "#B86092FF",  
  "Off-road machinery"         = "#00B7A7FF"   
)


# ---- Yearly shares by category ----
category_shares <- naei_totals |>
  summarise(emission = sum(benzene_emission, na.rm = TRUE),
            .by = c(year, final_category)) |>
  mutate(share = emission / sum(emission), .by = year) |>
  mutate(final_category = factor(final_category, levels = names(source_colours)))


# ---- Plot 1: share of emissions every year ----
p_share <- category_shares |>
  ggplot(aes(
    x = year, y = share, fill = final_category,
    text = paste0("<b>", final_category, "</b><br>",
                  "Year: ", year, "<br>",
                  "Share: ", percent(share, accuracy = 0.1), "<br>",
                  "Emissions: ", round(emission, 2), " kt")
  )) +
  geom_col(width = 0.85, colour = "white", linewidth = 0.2, alpha = 0.6) +  # thin white gaps between segments
  scale_fill_manual(values = source_colours, name = NULL) +
  scale_y_continuous(name = "Share of UK benzene emissions",
                     labels = percent, expand = c(0, 0), minor_breaks = seq(0,1,0.1), breaks = seq(0,1,0.2)) +
  scale_x_continuous(name = "Year", breaks = seq(1990, 2020, 5),
                     expand = c(0.01, 0)) +
  guides(fill = guide_legend(nrow = 2)) +
  theme_minimal(base_size) +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_line()
  )



show_plot(p_share)

```

However, a rising share does not mean rising emissions. Domestic emissions fell from 4.4 kt in 1990 to 2.6 kt in 2023; they simply fell more slowly than everything else. The next figures show the change in absolute terms. Within the domestic sector, the fuel mix has changed markedly. Coal burning, which produced 1.9 kt in 1990, now produces less than 0.1 kt. Over the same period, emissions from domestic wood burning have roughly tripled, from 0.5 kt to 1.7 kt, and wood is now the main source of domestic benzene. Waste and shipping have also become relatively more important as their emissions have changed little since 1990.


```{r relative-source}
#| echo: false
#| warning: false
#| fig-height: 6

# ---- Data: add % of 1990 to every row, used by both tabs ----
baseline_values <- category_shares |> 
  filter(year == 1990) |> 
  select(final_category, baseline_emission = emission)

category_trends <- category_shares |> 
  left_join(baseline_values, by = "final_category") |> 
  mutate(emissions_pct_baseline = emission / baseline_emission * 100)

# Same panel order in both tabs: biggest reduction first
panel_order <- category_trends |> 
  filter(year == max(year)) |> 
  arrange(emissions_pct_baseline) |> 
  pull(final_category)

category_trends <- category_trends |> 
  mutate(final_category = factor(final_category, levels = panel_order))

# ---- Shared style for both faceted charts ----
facet_style <- list(
  geom_line(linewidth = 0.8),
  geom_point(aes(text = paste0("<b>", final_category, "</b><br>",
                               "Year: ", year, "<br>",
                               "Emissions: ", round(emission, 2), " kt<br>",
                               "% of 1990: ", round(emissions_pct_baseline), "%")),
             size = 0.06),
  scale_colour_manual(values = source_colours, guide = "none"),
  scale_x_continuous(name = "Year",
                     breaks = seq(1990, 2020, 5),
                     labels = \(x) ifelse(x %in% c(1990, 2005, 2020), x, "")),
  theme_minimal(base_size),
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing.x = unit(1, "lines"),
    panel.spacing.y = unit(2, "lines"),
    strip.text = element_text(face = "bold", hjust = 0),
    axis.ticks.x = element_line(colour = "grey70"),
    axis.ticks.length.x = unit(4, "pt"),
    axis.line = element_line(colour = "grey80")
  )
)

# Shared plotly settings
show_facets <- function(p) {
  ggplotly(p, tooltip = "text", height = 550) |> 
    config(displayModeBar = FALSE)
}
```




::: {.panel-tabset}

## Absolute

Each panel shows benzene emissions from one source category in kilotonnes. *Note:* Every panel has a separately scaled y-axis.

```{r absolute-source}
#| label: fig-source-absolute
#| fig-cap: "Annual benzene emissions (kt) by source category, 1990–2023. Each panel has its own y-axis scale. Panels are ordered by the reduction since 1990, largest first."
#| echo: false
#| warning: false

p_absolute <- category_trends |> 
  ggplot(aes(x = year, y = emission, colour = final_category)) +
  facet_style +
  scale_y_continuous(name = "Emissions (kt)",
                     limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  facet_wrap(~ final_category, nrow = 2, scales = "free_y",
             labeller = label_wrap_gen(width = 18), axes = "all_x")

show_plot(p_absolute, height = 550)

```

## Relative to baseline

Each line shows a category's emissions as a percentage of the 1990 emissions. Road transport has fallen to 2% of its 1990 level, while waste (76%), shipping and other transport (69%) and domestic combustion (59%) have reduced by a lower fraction. These sources now make up a much bigger share of the total.

```{r}
#| label: fig-source-relative
#| fig-cap: "Benzene emissions by source category as a percentage of 1990 emissions. The dashed line marks the 1990 level (100%)."
#| echo: false
#| warning: false

p_relative <- category_trends |> 
  ggplot(aes(x = year, y = emissions_pct_baseline, colour = final_category)) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey70", linewidth = 0.4) +
  facet_style +
  scale_y_continuous(name = "Emissions (% of 1990)",
                     limits = c(0, 135), breaks = seq(0, 125, 25), expand = c(0, 0)) +
  facet_wrap(~ final_category, nrow = 2,
             labeller = label_wrap_gen(width = 18), axes = "all_x")

show_plot(p_relative, height = 550)
```

:::


These figures show that UK benzene emissions have changed from being dominated by a mobile distributed source to being spread across several smaller ones. Many of the remaining sources, such as refineries and industrial plants, are located in fixed places rather than spread along the road network. This means national totals may say less about local exposure than they once did, and it motivates looking at where emissions occur and what we are monitoring -  discussed in the next section.

### Emissions from Point Sources

Point sources are individual facilities, such as refineries, chemical works, steelworks and incinerators, whose emissions are reported to the NAEI at a specific location. Unlike diffuse sources like road traffic or domestic heating, their emissions are concentrated at one site, so they can have a disproportionate effect on air quality in the surrounding communities.

In 2005, point sources accounted for 10.6% of UK benzene emissions (1885 t). As of  2023, point sources accounted for 769 t, representing 9.6% of emissions. 

```{r point-sources-data}
#| echo: false
#| message: false
#| warning: false


# Created by scripts/point_sources_prep.R — rerun that script if the NAEI data changes
uk_outline_bng         <- read_rds("processed_data/uk_outline_bng.rds")
point_sources          <- read_rds("processed_data/point_sources.rds")
top_sites              <- read_rds("processed_data/top_sites.rds")
industrial_communities <- read_rds("processed_data/industrial_communities.rds")
 
# Interactive in HTML, static in Word/PDF 
if (!exists("show_plot")) {
  show_plot <- function(p, ...) {
    if (knitr::is_html_output()) {
      ggplotly(p, tooltip = "text", ...) |> config(displayModeBar = FALSE)
    } else {
      p
    }
  }
}
 
point_colour <- "#16317D"   # single colour for one-series charts
 
# ---- Key numbers used in the text ----
# point-source Emission is in tonnes; national totals are in kt
point_share <- point_sources |>
  st_drop_geometry() |>
  summarise(point_total_t = sum(Emission, na.rm = TRUE), .by = Year) |>
  left_join(yearly_emissions, by = c("Year" = "year")) |>
  mutate(national_t = total_emissions_kt * 1000,
         point_pct  = point_total_t / national_t * 100)
 
latest_year  <- max(point_share$Year)
latest_share <- point_share |> filter(Year == latest_year)
 
# Concentration: sort sites from largest to smallest
site_ranking <- point_sources |>
  st_drop_geometry() |>
  filter(Year == latest_year) |>
  summarise(Emission = sum(Emission, na.rm = TRUE), .by = c(Site, Sector)) |>
  arrange(desc(Emission)) |>
  mutate(
    rank           = row_number(),
    site_pct       = rank / n(),
    cumulative_pct = cumsum(Emission) / sum(Emission)
  )
 

n_sites     <- nrow(site_ranking)

top10_share <- site_ranking |> 
  filter(rank == 10) |> 
  pull(cumulative_pct)
```



```{r point-share}
#| label: fig-point-share
#| fig-cap: "Benzene emissions from point sources as a percentage of total UK emissions, 2005–2023. Source: NAEI point source database and national totals."
#| echo: false
#| warning: false
 
p_point_share <- point_share |>
  ggplot(aes(x = Year, y = point_pct)) +
  geom_line(colour = point_colour, linewidth = 0.8) +
  geom_point(aes(text = paste0("Year: ", Year, "<br>",
                               "Point sources: ", round(point_total_t), " t<br>",
                               "Share of UK total: ", round(point_pct, 1), "%")),
             colour = "black", size = 0.8) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = "Point sources (% of UK total)",
                     labels = \(x) paste0(x, "%"),
                     limits = c(0, 12.5), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal(base_size) +
  theme(panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "grey80"),
        axis.ticks = element_line(colour = "grey70"))
 
show_plot(p_point_share)
```
 
Total point-source emissions are dominated by a small number of sites. Of the `r n_sites` sites reporting benzene emissions in `r latest_year`, the ten largest account for `r scales::percent(top10_share, accuracy = 1)` of point-source emissions.
 
```{r point-concentration}
#| label: fig-point-concentration
#| fig-cap: "Cumulative share of point-source benzene emissions in 2023, with sites ordered from largest to smallest. A steep early rise means a small number of sites account for most emissions."
#| echo: false
#| warning: false
 
p_concentration <- site_ranking |>
  ggplot(aes(x = site_pct, y = cumulative_pct)) +
  geom_line(colour = point_colour, linewidth = 1) +
  geom_point(aes(text = paste0("<b>", Site, "</b><br>",
                               "Rank: ", rank, " of ", n_sites, "<br>",
                               "Emission: ", round(Emission, 4), " t<br>",
                               "Cumulative share: ", scales::percent(cumulative_pct, 0.1))),
             colour = "black", size = 0.3) +
  scale_x_continuous(name = "Sites, largest first (% of all sites)",
                     labels = scales::percent, breaks = seq(0, 1, 0.1),
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Cumulative share of point-source emissions",
                     labels = scales::percent, breaks = seq(0, 1, 0.1),
                     limits = c(0, 1.04), expand = c(0, 0)) +
  theme_minimal(base_size) +
  theme(panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "grey80"))
 
show_plot(p_concentration)
```
 
#### Largest point sources
 
```{r top-sites-map}
#| label: fig-top-sites-map
#| fig-cap: "Locations of the largest benzene point sources in 2023. Circle area is proportional to annual emissions (tonnes)."
#| echo: false
#| warning: false
#| column: page
 
if (knitr::is_html_output()) {
 
  # ---- Yearly emissions for each top site since 2005 ----
  site_trends <- point_sources |>
    st_drop_geometry() |>
    filter(Site %in% top_sites$Site, Year >= 2005) |>
    summarise(Emission = sum(Emission, na.rm = TRUE), .by = c(Site, Year)) |>
    arrange(Site, Year)
 
  # Main sector for each site (largest emitter in the latest year)
  site_sectors <- point_sources |>
    st_drop_geometry() |>
    filter(Site %in% top_sites$Site, Year == latest_year) |>
    summarise(Emission = sum(Emission, na.rm = TRUE), .by = c(Site, Sector)) |>
    slice_max(Emission, n = 1, by = Site, with_ties = FALSE) |>
    select(Site, Sector)
 
  # Tiny inline SVG line chart for the popup (no extra packages needed)
  sparkline_svg <- function(years, values, width = 180, height = 45,
                            colour = point_colour) {
    x <- scales::rescale(years, to = c(3, width - 3))
    y <- scales::rescale(values, to = c(height - 3, 3),
                         from = c(0, max(values, 1e-9)))
    pts <- paste(round(x, 1), round(y, 1), sep = ",", collapse = " ")
    sprintf(
      '<svg width="%d" height="%d" style="display:block;margin-top:4px">
         <line x1="0" y1="%d" x2="%d" y2="%d" stroke="#ddd"/>
         <polyline points="%s" fill="none" stroke="%s" stroke-width="2"/>
       </svg>',
      width, height, height - 3, width, height - 3, pts, colour
    )
  }
 
  # ---- Build the popup HTML for each site ----
  site_popups <- site_trends |>
    summarise(
      first_year      = min(Year),
      last_year       = max(Year),
      first_emission  = Emission[which.min(Year)],
      latest_emission = Emission[which.max(Year)],
      spark           = sparkline_svg(Year, Emission),
      .by = Site
    ) |>
    left_join(site_sectors, by = "Site") |>
    mutate(
      pct_change  = (latest_emission / first_emission - 1) * 100,
      change_text = case_when(
        first_emission == 0 ~ paste0("No emissions reported in ", first_year),
        pct_change < 0      ~ paste0("&#9660; ", abs(round(pct_change)), "% since ", first_year),
        TRUE                ~ paste0("&#9650; ", round(pct_change), "% since ", first_year)
      ),
      popup = paste0(
        "<div style='font-family:sans-serif;font-size:13px;line-height:1.4'>",
        "<b>", htmltools::htmlEscape(Site), "</b><br>",
        "<span style='color:#666'>", htmltools::htmlEscape(Sector), "</span><br>",
        "<b>", round(latest_emission, 1), " t</b> in ", last_year, "<br>",
        change_text,
        spark,
        "<div style='display:flex;justify-content:space-between;width:180px;",
        "font-size:11px;color:#888'><span>", first_year, "</span><span>",
        last_year, "</span></div>",
        "</div>"
      )
    ) |>
    select(Site, popup)
 
  # ---- Map ----
  top_sites |>
    left_join(site_popups, by = "Site") |>
    st_transform(4326) |>
    st_centroid() |>
    leaflet() |>
    addProviderTiles(providers$Esri.WorldGrayCanvas) |>   # no API key needed
    addCircleMarkers(
      radius = ~ scales::rescale(sqrt(Emission), to = c(5, 20)),
      color = point_colour, fillOpacity = 0.6, stroke = FALSE,
      label = ~ Site,                                      # hover: name only
      popup = ~ popup                                      # click: full details
    )
} else {
  ggplot() +
    geom_sf(data = uk_outline_bng, fill = "grey95", colour = "grey70") +
    geom_sf(data = top_sites, aes(size = Emission), colour = "darkblue",
             alpha = 0.6, show.legend = "point") +
    scale_size_area(name = "Emission (t/yr)", max_size = 10, breaks = c(1, 10, 50, 100)) +
    guides(colour = "none",
         size = guide_legend(override.aes = list(colour = "grey40"))) +
    theme_void(12)
}
```




## Monitoring

```{r monitoring-data}
#| echo: false
#| message: false
#| warning: false

# Created by R/prepare_monitoring_sites.R
monitoring_sites        <- read_rds("processed_data/monitoring_sites.rds")
sites_active_years      <- read_rds("processed_data/sites_active_years.rds")
nearest_monitor_by_year <- read_rds("processed_data/nearest_monitor_by_year.rds")
nearest_monitor_current <- read_rds("processed_data/nearest_monitor_current.rds")

# ---- Colours ----
status_colours <- c(active = "#4a3aa7", closed = "#b5b5b0")
source_marker  <- "#A40000"     # same red as "Industry" in the emissions charts

category_colours <- c(
  "Urban Traffic"       = "#2a78d6",
  "Urban Industrial"    = "#4a3aa7",
  "Urban Background"    = "#1baf7a",
  "Suburban Background" = "#eda100",
  "Rural Background"    = "#e34948"
)

# ---- Numbers for the text ----
n_active      <- sum(monitoring_sites$status == "active")
n_active_ahn  <- sum(monitoring_sites$status == "active" & monitoring_sites$network == "AHN")
n_active_nahn <- sum(monitoring_sites$status == "active" & monitoring_sites$network == "NAHN")
n_ever        <- nrow(monitoring_sites)

sites_per_year <- sites_active_years |>
  st_drop_geometry() |>
  count(year, network_label, `Environment Type`, name = "n_sites") |>
  mutate(`Environment Type` = factor(`Environment Type`, levels = names(category_colours)))

peak <- sites_per_year |>
  summarise(n_sites = sum(n_sites), .by = year) |>
  slice_max(n_sites, n = 1, with_ties = FALSE)
```

Benzene in ambient air is regulated under the Air Quality Standards Regulations 2010, which set an annual mean limit value of 5 µg m^-3^. Defra monitors it through the UK Hydrocarbons Network, which is made up of two complementary parts: a small number of automatic sites giving high time-resolution data, and a larger network of non-automatic samplers giving wider spatial coverage. In 2023, no UK monitoring site exceeded the limit value.

### Automatic Hydrocarbon Network

The Automatic Hydrocarbon Network uses automated gas chromatography to measure 29 volatile organic compounds, including benzene and 1,3-butadiene, every hour. It currently has `r n_active_ahn` active sites. The hourly data are valuable for understanding daily cycles and short pollution episodes and for identifying sources. However, with so few sites, the network cannot say much about how concentrations vary across the country.

### Non-Automatic Hydrocarbon Network

The Non-Automatic Hydrocarbon Network gives much wider coverage, with `r n_active_nahn` active sites. Air is drawn through dual sorbent tubes by a controlled-flow pump. The tubes are changed every two weeks and analysed in the laboratory by thermal desorption and gas chromatography–mass spectrometry. This gives fortnightly averages, not hourly values, but the extra sites cover a far wider range of locations.

### Network coverage over time

`r n_ever` sites have measured benzene in the two networks at some point, and `r n_active` are active today. Coverage peaked in `r peak$year`, with `r peak$n_sites` sites running.

::: {.panel-tabset}

## Sites per year

```{r sites-per-year}
#| label: fig-sites-per-year
#| fig-cap: "Number of benzene monitoring sites operating each year across the automatic and non-automatic networks, by site type. A site is counted if it operated for any part of the year. One site classified as 'Unknown Industrial' (2008–2010) is shown as Urban Industrial."
#| echo: false
#| warning: false

p_sites_year <- sites_per_year |>
  summarise(n_sites = sum(n_sites), .by = c(year, `Environment Type`)) |>
  ggplot(aes(x = year, y = n_sites, fill = `Environment Type`,
             text = paste0("<b>", `Environment Type`, "</b><br>",
                           "Year: ", year, "<br>",
                           "Sites: ", n_sites))) +
  geom_col(width = 0.85, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = category_colours, name = NULL) +
  scale_y_continuous(name = "Number of sites", expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(name = NULL,
                     breaks = seq(1995, 2025, 5),
                     labels = \(x) ifelse(x %% 10 == 0, x, "")) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(caption = "One site classified as \"Unknown Industrial\" (2008–2010) is shown as Urban Industrial.") +
  theme_minimal(base_size) +
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "grey80"),
    axis.ticks.x = element_line(colour = "grey70")
  )

show_plot(p_sites_year)
```

## Active and closed sites

```{r site-status}
#| label: fig-site-status
#| fig-cap: "Active and closed benzene monitoring sites in each network, by site type."
#| echo: false
#| warning: false

p_status <- monitoring_sites |>
  st_drop_geometry() |>
  count(network_label, status, `Environment Type`, name = "count") |>
  mutate(`Environment Type` = fct_reorder(`Environment Type`, count, sum),
         status = factor(status, levels = c("closed", "active"))) |>
  ggplot(aes(x = count, y = `Environment Type`, fill = status,
             text = paste0("<b>", `Environment Type`, "</b><br>",
                           network_label, ", ", status, ": ", count, " sites"))) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.2) +
  facet_wrap(~ network_label, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = status_colours, name = NULL,
                    breaks = c("active", "closed"),
                    labels = c("Active", "Closed")) +
  scale_x_continuous(name = "Number of monitoring sites",
                     expand = expansion(mult = c(0, 0.05))) +
  labs(y = NULL) +
  theme_minimal(base_size) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0)
  )

show_plot(p_status)
```

:::

## Location of Emissions relative to Monitoring

```{r location-numbers}
#| echo: false

distance_stats <- nearest_monitor_by_year |>
  summarise(
    median_km   = median(distance_km, na.rm = TRUE),
    weighted_km = weighted.mean(distance_km, w = Emission, na.rm = TRUE),
    .by = Year
  ) |>
  arrange(Year)

n_communities      <- nrow(nearest_monitor_current)
n_no_monitor_10km  <- sum(nearest_monitor_current$monitors_within_10km == 0)
median_current_km  <- median(nearest_monitor_current$distance_km)
```

The monitoring networks were largely designed when road transport dominated benzene emissions, so most sites are in urban areas. As the emission profile has shifted towards domestic combustion and industrial point sources, it is worth asking whether the networks still sample the places where people are most exposed.

Of the `r n_communities` largest point sources in `r latest_year`, `r n_no_monitor_10km` have no active benzene monitor within 10 km. The median distance from one of these sites to its nearest active monitor is `r round(median_current_km)` km.

```{r monitoring-map}
#| label: fig-monitoring-map
#| fig-cap: "Active benzene monitoring sites and the ten largest point sources in 2023 (red), each shown with a 10 km radius."
#| echo: false
#| warning: false
#| column: page

active_sites <- monitoring_sites |> filter(status == "active")
closed_sites <- monitoring_sites |> filter(status == "closed")

if (knitr::is_html_output()) {

  # Reuse the point-source popups if the earlier chunk created them
  map_sources <- top_sites
  if (exists("site_popups")) {
    map_sources <- map_sources |> left_join(site_popups, by = "Site")
  } else {
    map_sources <- map_sources |> mutate(popup = Site)
  }
  map_sources <- map_sources |> st_transform(4326) |> st_centroid()

  site_label <- \(d) paste0(d$`Site Name`, " (", d$network_label, ", ",
                            d$`Environment Type`, ")")

  leaflet() |>
    addProviderTiles(providers$Esri.WorldGrayCanvas, group = "Grey") |>
    addProviderTiles(providers$Esri.WorldImagery,    group = "Satellite") |>
    # 10 km industrial-community radius
    addCircles(data = map_sources, radius = 10000,
               color = source_marker, weight = 1, fillOpacity = 0.05,
               group = "10 km around major sources") |>
    addCircleMarkers(data = st_transform(closed_sites, 4326),
                     radius = 4, color = status_colours[["closed"]],
                     fillOpacity = 0.8, stroke = FALSE,
                     label = site_label(closed_sites),
                     group = "Closed monitoring sites") |>
    addCircleMarkers(data = st_transform(active_sites, 4326),
                     radius = 5, color = status_colours[["active"]],
                     fillOpacity = 0.9, stroke = FALSE,
                     label = site_label(active_sites),
                     group = "Active monitoring sites") |>
    addCircleMarkers(data = map_sources,
                     radius = 7, color = source_marker, weight = 2,
                     fillOpacity = 0.3,
                     label = ~ Site, popup = ~ popup,
                     group = "Major point sources") |>
    addLayersControl(
      baseGroups    = c("Grey", "Satellite"),
      overlayGroups = c("Active monitoring sites", "Closed monitoring sites",
                        "Major point sources", "10 km around major sources"),
      options = layersControlOptions(collapsed = FALSE)
    ) |>
    hideGroup("Closed monitoring sites") |>
    addLegend(position = "bottomright",
              colors = c(status_colours[["active"]], status_colours[["closed"]], source_marker),
              labels = c("Active monitor", "Closed monitor", "Major point source"),
              opacity = 0.9)

} else {
  ggplot() +
    geom_sf(data = uk_outline_bng, fill = "grey95", colour = "grey70") +
    geom_sf(data = st_buffer(top_sites, 10000),
            fill = source_marker, alpha = 0.08, colour = source_marker, linewidth = 0.2) +
    geom_sf(data = active_sites, aes(shape = network_label),
            colour = status_colours[["active"]], size = 2) +
    geom_sf(data = top_sites, colour = source_marker, shape = 4, size = 3, stroke = 1.2) +
    scale_shape_manual(values = c(Automatic = 17, `Non-automatic` = 16), name = "Active monitor") +
    labs(caption = "Red crosses: largest point sources, with 10 km radius.") +
    theme_void(12) +
    theme(legend.position = "top")
}
```

#### Nearest monitor to each major source

```{r nearest-table}
#| label: tbl-nearest-monitor
#| tbl-cap: "Nearest active benzene monitor to each of the ten largest point sources in 2023, and the number of active monitors within 10 km."
#| echo: false

nearest_monitor_current |>
  transmute(
    `Point source`       = Site,
    `Emission (t)`       = round(Emission, 1),
    `Nearest monitor`    = paste0(nearest_site, " (", nearest_network, ")"),
    `Monitor type`       = nearest_type,
    `Distance (km)`      = round(distance_km, 1),
    `Monitors within 10 km` = monitors_within_10km
  ) |>
  knitr::kable(align = "lrllrr")
```

#### Has monitoring kept up with the sources?

The chart below compares each year's 100 largest point sources with the monitors running in that year. It shows the median distance to the nearest monitor, and the mean weighted by emissions, which gives more influence to the largest sources.

```{r distance-trend}
#| label: fig-distance-trend
#| fig-cap: "Distance from each year's 100 largest point sources to the nearest monitor operating in that year. The emission-weighted mean gives more influence to the largest sources."
#| echo: false
#| warning: false

p_distance <- distance_stats |>
  pivot_longer(c(median_km, weighted_km), names_to = "measure", values_to = "km") |>
  mutate(measure = recode(measure,
                          median_km   = "Median",
                          weighted_km = "Emission-weighted mean")) |>
  ggplot(aes(x = Year, y = km, linetype = measure)) +
  geom_line(colour = point_colour, linewidth = 0.8) +
  geom_point(aes(text = paste0("<b>", measure, "</b><br>",
                               "Year: ", Year, "<br>",
                               "Distance: ", round(km, 1), " km")),
             colour = point_colour, size = 0.6) +
  scale_linetype_manual(values = c("Median" = "solid",
                                   "Emission-weighted mean" = "dashed"),
                        name = NULL) +
  scale_y_continuous(name = "Distance to nearest active monitor (km)",
                     limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(name = NULL) +
  theme_minimal(base_size) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "grey80"),
    axis.ticks = element_line(colour = "grey70")
  )

show_plot(p_distance)
```



### Comparing automatic and non-automatic measurements
 
```{r comparison-data}
#| echo: false
#| message: false
 
# Created by R/prepare_network_comparison.R
closest_pairs          <- read_rds("processed_data/closest_network_pairs.rds")
comparison_sites       <- read_rds("processed_data/comparison_sites.rds")
comparison_fortnightly <- read_rds("processed_data/comparison_fortnightly.rds")
 
# One colour per site, used in both the map and the plot
comparison_colours <- c(
  "London Marylebone Road" = "#2a78d6",
  "Camden Kerbside"        = "#eb6834",
  "London Bloomsbury"      = "#1baf7a"
)
```
 
No automatic and non-automatic monitors share a location, so the two methods cannot be compared directly. The table below lists the ten closest pairs of automatic and non-automatic sites that operated in the same years.
 
```{r closest-pairs-table}
#| label: tbl-closest-pairs
#| tbl-cap: "The ten closest pairs of automatic and non-automatic monitoring sites that operated in the same years."
#| echo: false
 
closest_pairs |>
  transmute(
    `Automatic site`         = ahn_site,
    `Type`                   = ahn_type,
    `Non-automatic site`     = nahn_site,
    `Type `                  = nahn_type,
    `Distance (km)`          = round(distance_km, 1),
    `Years operating together` = paste0(first_year, "–", last_year,
                                        " (", overlap_years, " yrs)")
  ) |>
  knitr::kable(align = "llllrl")
```
 
London Marylebone Road (automatic) was compared with two nearby non-automatic sites: Camden Kerbside, another roadside site, and London Bloomsbury, an urban background site.
 
```{r comparison-map}
#| label: fig-comparison-map
#| fig-cap: "Locations of London Marylebone Road (automatic), Camden Kerbside and London Bloomsbury (non-automatic)."
#| echo: false
#| warning: false
 
sites_4326 <- comparison_sites |> st_transform(4326)
ahn_point  <- sites_4326 |> filter(network_label == "Automatic")
nahn_pts   <- sites_4326 |> filter(network_label != "Automatic")
 
# Distances from Marylebone Road for the labels
nahn_pts <- nahn_pts |>
  mutate(distance_km = as.numeric(st_distance(geometry, ahn_point$geometry)) / 1000)
 
if (knitr::is_html_output()) {
 
  # Lines from Marylebone Road to each non-automatic site
  links <- st_sfc(
    lapply(seq_len(nrow(nahn_pts)), \(k)
      st_linestring(rbind(st_coordinates(ahn_point), st_coordinates(nahn_pts[k, ])))),
    crs = 4326
  )
 
  leaflet() |>
   addProviderTiles(providers$Esri.WorldGrayCanvas) |>
    addPolylines(data = links, color = "grey40", weight = 1.5, dashArray = "4",
                 label = paste0(round(nahn_pts$distance_km, 1), " km")) |>
    addCircleMarkers(
      data = sites_4326,
      radius = 8, stroke = TRUE, color = "white", weight = 2, fillOpacity = 1,
      fillColor = ~ unname(comparison_colours[site]),
      label = ~ paste0(site, " (", network_label, ", ", `Environment Type`, ")"),
      labelOptions = labelOptions(noHide = TRUE, direction = "right")
    )
 
} else {
  ggplot() +
    ggspatial::annotation_map_tile(type = "osm", zoom = 14, progress = "none") +
    ggspatial::annotation_scale(
    location = "bl",
    width_hint = 0.25,
    text_cex = 0.8
  ) +
    geom_sf(data = sites_4326, aes(colour = site), size = 4) +
    geom_sf_label(data = sites_4326, aes(label = site), nudge_y = 0.0015,
                  size = 3, label.size = 0) +
    scale_colour_manual(values = comparison_colours, guide = "none") +
    coord_sf(expand = TRUE) +
    theme_void()
}
```
 
The chart below shows fortnightly benzene concentrations at the three sites. The automatic values are hourly measurements averaged over the same two-week periods as the non-automatic samples.
 
```{r comparison-plot}
#| label: fig-comparison-fortnightly
#| fig-cap: "Fortnightly benzene concentrations at London Marylebone Road (automatic, hourly data averaged over each sampling period), Camden Kerbside and London Bloomsbury (non-automatic). Gaps of more than one month are left unjoined."
#| echo: false
#| warning: false
#| column: page
 
p_comparison <- comparison_fortnightly |>
  ggplot(aes(x = mid_date, y = benzene, colour = site)) +
  geom_line(aes(group = interaction(site, segment)), linewidth = 0.5, alpha = 0.8) +
  geom_point(aes(text = paste0("<b>", label, "</b><br>",
                               format(start_date, "%d %b %Y"), " to ",
                               format(end_date, "%d %b %Y"), "<br>",
                               "Benzene: ", round(benzene, 2), " µg m⁻³")),
             size = 0.8) +
  scale_colour_manual(values = comparison_colours, name = NULL,
                      labels = c("London Marylebone Road" = "London Marylebone Road (automatic)",
                                 "Camden Kerbside(Swiss Cottage)"        = "Camden Kerbside (non-automatic)",
                                 "London Bloomsbury"      = "London Bloomsbury (non-automatic)")) +
  scale_y_continuous(name = "Benzene (µg m⁻³)", limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_x_date(name = NULL) +
  theme_minimal(base_size) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "grey80"),
        axis.ticks.x = element_line(colour = "grey70"))
 
show_plot(p_comparison)
```