---
title: "Industrial Case Studies"
format: 
  html:
    toc: true
editor: source
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| message: false
#| warning: false
#| echo: false


knitr::opts_knit$set(root.dir = "G:/My Drive/Year 3/Industrial Communities/")
library(tidyverse)
library(plotly)
library(sf)
library(leaflet)
library(terra)
library(IMD)
library(sfarrow)


```




## Industrial Communities

As the UK transitions to Net Zero, there becomes a concentration of industrial activities in certain areas. There is a risk that, in the reduction of pollutant concentration levels below regulatory limits, monitoring is reduced, or even stopped, as the case may be for Benzene. Here, we take a closer look into several industrial areas remaining in the UK that remain large point sources of Benzene, the nature of the industrial processes ongoing (and how this may change in a Net Zero world) as well as the demographic make up. 


## Total Benzene Emissions 


```{r}


naei_totals_raw <- read_csv("data/naei_inventory_total/naei_uk_emissions_data_benzene.csv")

point_sources <- read_rds("processed_data/point_sources.rds")

NFR_codes <- read_csv("data/naei_inventory_total/NFRcodes_descriptions.csv")

naei_totals <- naei_totals_raw |> 
  pivot_longer(cols = c(`1990`:`2023`), names_to = "year", values_to = "benzene_emission") |> 
  mutate(benzene_emission = as.numeric(benzene_emission), 
         year = as.numeric(year)) |> 
  left_join(NFR_codes, join_by(`NFR/CRT Group` == `NFR Code`))

ggplotly(
ggplot(naei_totals) +
  geom_col(aes(x = year, y = benzene_emission, fill = Source)) +
  scale_y_continuous(name = "Emission (kilotonnes)") +
  scale_x_continuous(name = "Year") +
  theme_minimal() +
  theme(legend.position = "none")
)


# And a simple line graph....


totals_by_year <- naei_totals |> 
  group_by(year) |> 
  summarise(yearly_totals = sum(benzene_emission, na.rm = T))


total_1990 <- totals_by_year |> 
  filter(year == 1990) |> 
  select(yearly_totals) |> 
  pull()

ggplotly(
  totals_by_year |> 
     mutate(rel_to_1990 = yearly_totals/total_1990 * 100) |> 
    mutate(tooltip_text = paste0(
      "Year: ", year, "\n",
      "Emission: ", round(yearly_totals, 2), " kt\n",
      "Relative to 1990: ", round(rel_to_1990, 1), "%"
    )) |> 
  ggplot(aes(x = year, y = yearly_totals, text = tooltip_text)) +

  geom_point(colour = "#8A7090") +
  geom_line(group = 1) +
  scale_y_continuous(name = "Emission (kilotonnes)", limits = c(0,NA)) +
  scale_x_continuous(name = "Year") +
  theme_minimal(), 
  tooltip = "text"
)



```

```{r}
point_source_map <- function(point_sources_df = point_sources,
                           centre_lng,
                           centre_lat,
                           year = 2023,
                           distance = 15000,
                           zoom = 11) {

  # Create centre point
  centre_point <- data.frame(
    lng = centre_lng,
    lat = centre_lat
  ) |>
    st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
    st_transform(27700)

  # Filter points
  map_points <- point_sources_df |>
    filter(
      drop(
        st_is_within_distance(
          geometry,
          centre_point,
          dist = distance,
          sparse = FALSE
        )
      ),
      Year == year
    ) |>
    st_transform(4326)

  # Colour palette
  sector_palette <- colorFactor(
    palette = "Set2",
    domain = map_points$Sector
  )

  # Build map
  leaflet(map_points) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    setView(
      lng = centre_lng,
      lat = centre_lat,
      zoom = zoom
    ) |>
    addCircleMarkers(
      radius = ~scales::rescale(
        sqrt(Emission),
        to = c(5, 18)
      ),
      stroke = TRUE,
      weight = 1,
      color = "white",
      fillColor = ~sector_palette(Sector),
      fillOpacity = 0.8,
      popup = ~paste0(
        "<b>", Site, "</b><br>",
        "<b>Sector:</b> ", Sector, "<br>",
        "<b>Year:</b> ", year, "<br>",
        scales::comma(Emission, accuracy = 0.01),
        " kg"
      )
    ) |>
    addLegend(
      title = htmltools::HTML("<strong>Industry sector</strong>"),
      pal = sector_palette,
      values = ~Sector,
      position = "bottomleft",
      opacity = 0.8
    ) |>
    addScaleBar(position = "bottomright") |> 
    addEasyButton(easyButton(
      icon = "fa-list",
      title = "Show/hide legend",
      onClick = htmlwidgets::JS(" function(btn,map){ $('.legend').toggle(); } ")))
  
  
  
}





point_sources_timeline <- function(point_sources_df = point_sources, 
                                   centre_lat, 
                                   centre_lng, 
                                   distance = 15000) {
  
  
  # Create centre point
  centre_point <- data.frame(
    lng = centre_lng,
    lat = centre_lat
  ) |>
    st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
    st_transform(27700)
  
  
  # Filter dataset to be within this radius.... 
  
  
  
  # Working out a threshold at which we keep the data... 
  # if it is more than 1% of the total
  
  
  filtered_data <- point_sources_df |> 
     filter(
      drop(
        st_is_within_distance(
          geometry,
          centre_point,
          dist = distance,
          sparse = FALSE
        )
      )) |> 
    filter(Emission >= quantile(Emission, 0.01))
  
  
  
 filtered_data|> 
    ggplot(aes(x = Year, y = Emission, colour = Sector, group = Site), group = 1) +
    geom_point() +
    geom_line() +
    #scale_colour_discrete(palette = "Set2") +
    facet_wrap(~Sector, scales = "free_y") +
    theme(legend.position = "none")
  
  
}




```



### Humberside



```{r}
# Lets get a plot of the area showing the main benzene sources.... 


humberside_map <- point_source_map(centre_lat = 53.6545, centre_lng = -0.2310)

humberside_map

ggplotly(
point_sources_timeline(centre_lat = 53.6545, centre_lng = -0.2310)
)
```


Key activities that are going on.....

| Facility | Operator | Type | Scale |
|-----------|----------|------|-------|
| South Killingholme Refinery | Phillips 66 | Processing and distribution of petroleum products | Large |
| Lindsey Oil Refinery | Prax | Processing and distribution of petroleum products | Large |
| Immingham Docks | Associated British Ports | Shipping and logistics | Large |

### Teeside

```{r}
teeside_map <- point_source_map(centre_lat = 54.610, centre_lng = -1.168)

teeside_map

point_sources_timeline(centre_lat = 54.610, centre_lng = -1.168)
```



### Milford Haven 


```{r}
milford_haven_map <- point_source_map(centre_lat = 51.710, centre_lng = -5.037)

milford_haven_map

point_sources_timeline(centre_lat = 51.710, centre_lng = -5.037)
```


### Stanlow

```{r}
stanlow_map <- point_source_map(centre_lat = 53.2738, centre_lng = -2.84)

stanlow_map 
```


### Fawley

```{r}
fawley_map <- point_source_map(centre_lat =50.84, centre_lng = -1.362)

fawley_map 
```

### Grangemouth

```{r}
grangemouth_map <- point_source_map(centre_lat =56.020, centre_lng = -3.694)


grangemouth_map 
```

## Deprivation and Environmental Justice

As the point-source maps above show, several of the UK's largest remaining benzene emitters sit close to residential areas. This section asks a more formal question: are the communities living near these sites more deprived than the national average, and does that relationship track the scale of emissions?

We use the [mySociety composite UK Index of Multiple Deprivation](https://github.com/mysociety/composite_uk_imd) (2020) via the `IMD` R package, which rescales England's, Wales's, and Scotland's separate indices onto a single comparable UK-wide scale (population-weighted rank standardisation) — this matters here because our sites span all three nations, and their *native* indices (English IMD 2025, WIMD, SIMD 2020) are not directly comparable to one another. Population weighting within each buffer uses the Carnell et al. 1km² gridded population density product, rather than simple LSOA/Data Zone counts, so that exposure estimates reflect where people actually live within an area rather than assuming they're evenly spread across it.

::: {.callout-note}
**A note on buffer distance.** Rather than picking one arbitrary radius, we use a set of **concentric distance bands (1, 3, 5, and 10 km)** around each site, with **5 km treated as the primary "impact zone"** for headline comparisons (RQ1, RQ3, RQ5) and the full gradient used for the dose-response question (RQ2).

The reasoning: benzene is a reactive, buoyant VOC — concentrations from an elevated point source (refinery stacks, tank farms) typically fall off quickly with distance, with most epidemiological and environmental-justice studies of refinery-adjacent communities using buffers somewhere in the 1–10 km range rather than the 15 km used for the map visuals above (which was chosen for cartographic context, not exposure estimation). Using bands instead of a single cut-off avoids baking an assumption about dispersion distance into the result, and directly supports RQ2 (does deprivation increase as you get closer to source?). If a follow-up phase adds dispersion modelling or prevailing wind direction, these bands are also a natural scaffold to refine into a directional/wind-weighted exposure surface later.
:::

### Setup: deprivation and population data

```{r}
#| label: deprivation-setup

library(IMD)     # humaniverse composite UK IMD

# --- 1. Composite UK IMD (population-weighted, comparable across nations) ---
# load_composite_imd() returns one nation's boundary/code set at a time;
# we load all three nations covered by our sites and bind them together.
imd_uk <- load_composite_imd(nation = "E")


# --- 2. Small-area boundaries (needed to spatially join sites to IMD) ---
# The IMD package supplies *scores*, not geometries, so LSOA (England/Wales)
# and Data Zone (Scotland) boundary polygons need to be sourced separately,
# e.g. from the ONS Open Geography Portal / NRS, and read in here as `sf`
# objects with the same area-code column used in imd_uk.
#

small_area_boundaries <- st_read_parquet("data/geography/uk-small-area-lsoa-soa-dz-apr-2021.parquet") |> 
  rename(lsoa = areacode)


# --- 3. Carnell et al. 1km population density grid ---
# Adjust path/column names to match your saved copy.
pop_grid <- rast("data/geography/population_per_1km2/data/uk_residential_population_2021.tif") |>
  as.points() |> 
  st_as_sf() |> 
  st_transform(27700)
# Expected columns (adjust as needed): geometry (1km cell, point or polygon
# centroid) and a population/population-density field, e.g. `population`.

# --- 4. Site list (reuse the coordinates already defined for the maps) ---
sites <- tibble::tribble(
  ~site,            ~centre_lat, ~centre_lng, ~nation,
  "Humberside",      53.6545,    -0.2310,     "England",
  "Teesside",        54.610,     -1.168,      "England",
  "Milford Haven",   51.710,     -5.037,      "Wales",
  "Stanlow",         53.2738,    -2.84,       "England",
  "Fawley",          50.84,      -1.362,      "England",
  "Grangemouth",     56.020,     -3.694,      "Scotland"
) |>
  st_as_sf(coords = c("centre_lng", "centre_lat"), crs = 4326, remove = FALSE) |>
  st_transform(27700)

buffer_distances <- c(1000, 3000, 5000, 10000)  # metres
```

### Population-weighted deprivation exposure

```{r}
#| label: pop-weighted-exposure

# Computes the population-weighted mean UK-composite IMD score (and decile)
# for a given site, at a given buffer distance, using the population grid
# joined to small-area IMD scores.
#
# NB: this assumes `small_area_boundaries` (with area-code + geometry) and
# `pop_grid` (with population + geometry) are loaded as above — swap in the
# real object/column names once those datasets are read in.

calc_site_exposure_by_decile <- function(site_geom, distance, boundaries, pop_grid, imd_lookup) {

  buffer <- st_buffer(site_geom, dist = distance)

  # Population grid cells falling inside the buffer
  cells_in_buffer <- pop_grid[st_within(pop_grid, buffer, sparse = FALSE)[, 1], ]

  # Assign each cell to its small area, then attach that area's IMD score
  cells_with_imd <- cells_in_buffer |>
    st_join(boundaries) |>
    left_join(imd_lookup, by = "lsoa")  # match join key to your data

  # Population-weighted mean deprivation score
   cells_with_imd |>
    st_drop_geometry() |>
    group_by(lsoa, E_expanded_decile) |>
    summarise(population = sum(pop_aw, na.rm = TRUE), .groups = "drop")
}

# Example of how this scales across all sites x all buffer distances once
# the input datasets are wired up:
#
exposure_by_decile <- sites |>
  tidyr::crossing(distance = buffer_distances) |>
  rowwise() |>
  mutate(
    lsoa_exposure = list(calc_site_exposure_by_decile(geometry, distance,
                                                        small_area_boundaries, pop_grid, imd_uk))
  ) |>
  tidyr::unnest(lsoa_exposure) |>
  group_by(site, distance, E_expanded_decile) |>
  summarise(population = sum(population), .groups = "drop")
```

### Q1 — Are point-source communities more deprived than the national average?

Compare each site's population-weighted mean IMD score/decile (5 km band) against the UK national median decile (5.5), e.g. with a one-sample Wilcoxon test, and visualise as a bar chart of "% of local population in the most deprived national decile" vs the 10% expected nationally.

```{r}
#| label: rq1-deprivation-vs-national

exposure_by_decile |>
  filter(distance == 3000) |>
  mutate(E_expanded_decile = factor(E_expanded_decile, levels = 1:10)) |>
  ggplot(aes(x = reorder(site, -population), y = population, fill = E_expanded_decile)) +
  geom_col() +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, name = "IMD decile\n(1 = most deprived)") 
 



# 
# exposure_results |>
#   filter(distance == 5000) |>
#   ggplot(aes(x = reorder(site, pop_weighted_score), y = pop_weighted_score)) +
#   geom_point(colour = "#8A7090") +
#   geom_hline(yintercept = 5, linetype = "dashed") +
#   coord_flip() +
#   scale_y_continuous(limits = c(1,10), breaks = seq(1,10,1), name = "Population-weighted UK composite IMD decile (5km)") +
#   theme_minimal()
# 
# 
# 
# exposure_results |>
#   filter(distance == 3000) |>
#   ggplot(aes(x = reorder(site, pop_weighted_score), y = total_population)) +
#   geom_col(colour = "#8A7090") +
#   geom_hline(yintercept = 5, linetype = "dashed") +
#   coord_flip() +
#   #scale_y_continuous(limits = c(1,10), breaks = seq(1,10,1), name = "Population-weighted UK composite IMD decile (5km)") +
#   theme_minimal()

```

### Q2 — Dose-response: does deprivation change with distance from source?

Use the full `buffer_distances` gradient to test whether deprivation score decreases as buffer size increases (i.e. the closest ring is most deprived), via Spearman correlation between distance band and population-weighted score, per site and pooled.

```{r}
#| label: rq2-dose-response

# exposure_results |>
#   ggplot(aes(x = distance / 1000, y = pop_weighted_score, colour = site)) +
#   geom_line() +
#   geom_point() +
#   scale_x_continuous(name = "Buffer radius (km)") +
#   scale_y_continuous(name = "Population-weighted IMD score", limits = c(0,10)) +
#   theme_minimal()
```

### Q3 — Which deprivation domains matter?

Repeat the RQ1 comparison using domain-level scores (income, employment, health, education, living environment, etc.) rather than the composite, to see whether the pattern is driven by economic deprivation, health deprivation, or something else. Note: the Living Environment domain can partly encode local environmental quality itself, so treat any signal there with caution to avoid circularity with the emissions data.

### Q4 — Has deprivation near these sites changed as emissions have fallen?

IMD releases are not annual (2004/2007/2010/2015/2019/2025 for England; similar sparse schedules for Wales/Scotland), so this can't be matched to the continuous NAEI time series above. Instead, compare the same LSOAs/Data Zones across two available IMD releases (e.g. 2019 vs 2025 composite equivalents) against the emissions trend over the same window, and report qualitatively rather than as a continuous regression.

### Q5 — How do the six sites compare to each other?

Rank sites by 5 km population-weighted exposure score and total exposed population, and check whether cross-nation comparability (via the composite index) changes the ranking compared with using each site's native national index.

### Limitations

- Point-source emissions are a proxy for exposure, not a measured ambient concentration — without dispersion modelling this analysis cannot say how much benzene any individual actually breathes.
- IMD (and the mySociety composite) measures deprivation of small *areas*, not individuals — results describe who lives nearby, not who is personally deprived.
- The composite UK IMD trades some national-methodology precision for cross-country comparability; where within-country conclusions matter most, cross-check against each nation's native index.
- IMD snapshots are infrequent relative to the annual emissions series, limiting how precisely temporal claims (RQ4) can be made.


```{r}
#| label: lsoa-pop-density-map

library(leaflet.extras)  # for addHeatmap()

industrial_deprivation_map <- function(point_sources_df = point_sources,
                                       centre_lng,
                                       centre_lat,
                                       boundaries,
                                       pop_grid,
                                       imd_lookup,
                                       year = 2023,
                                       distance = 15000,
                                       decile_col = "E_expanded_decile",
                                       zoom = 11) {
  

  # --- Shared centre point / buffer ---
  centre_point <- data.frame(lng = centre_lng, lat = centre_lat) |>
    st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
    st_transform(27700)
  
  buffer <- st_buffer(centre_point, dist = distance)
  
  # --- Layer 1: point sources ---
  map_points <- point_sources_df |>
    filter(
      drop(st_is_within_distance(geometry, centre_point, dist = distance, sparse = FALSE)),
      Year == year
    ) |>
    st_transform(4326)
  
  sector_palette <- colorFactor(palette = "Set2", domain = map_points$Sector)
  
  # --- Layer 2: LSOAs / IMD decile ---
  lsoas_in_buffer <- boundaries |>
    st_filter(buffer, .predicate = st_intersects) |>
    left_join(imd_lookup, by = "lsoa") |>
    st_transform(4326)
  
  decile_pal <- colorFactor(palette = "RdYlBu", domain = factor(1:10), reverse = TRUE)
  
  # --- Layer 3: population density heatmap ---
  pop_in_buffer <- pop_grid[
    st_within(pop_grid, st_transform(buffer, st_crs(pop_grid)), sparse = FALSE)[, 1],
  ] |>
    st_transform(4326)
  
  pop_coords <- pop_in_buffer |>
    st_drop_geometry() |>
    bind_cols(st_coordinates(pop_in_buffer) |> as.data.frame() |> rename(lng = X, lat = Y))
  
  # --- Combined map ---
  leaflet(options = leafletOptions(preferCanvas = TRUE)) |>
    addProviderTiles(providers$CartoDB.Positron) |>
    setView(lng = centre_lng, lat = centre_lat, zoom = zoom) |>
    
    # bottom: deprivation choropleth
    addPolygons(
      data = lsoas_in_buffer,
      fillColor = ~decile_pal(factor(get(decile_col))),
      fillOpacity = 0.5,
      color = "white",
      weight = 0.5,
      label = ~paste0("LSOA: ", lsoa, " | Decile: ", get(decile_col)),
      group = "IMD decile"
    ) |>
    
    # middle: population heatmap
    addHeatmap(
      data = pop_coords,
      lng = ~lng, lat = ~lat,
      intensity = ~pop_aw,
      radius = 22, blur = 28,
      max = max(log1p(pop_coords$pop_aw)),
      gradient = "magma",
      group = "Population density"
    ) |>
    
    # top: point sources
    addCircleMarkers(
      data = map_points,
      radius = ~scales::rescale(sqrt(Emission), to = c(5, 18)),
      stroke = TRUE, weight = 1, color = "white",
      fillColor = ~sector_palette(Sector),
      fillOpacity = 0.85,
      popup = ~paste0(
        "<b>", Site, "</b><br>",
        "<b>Sector:</b> ", Sector, "<br>",
        "<b>Year:</b> ", year, "<br>",
        scales::comma(Emission, accuracy = 0.01), " kg"
      ),
      group = "Point sources"
    ) |>
    
    # legends — separate corners so they don't overlap
    addLegend(
      title = htmltools::HTML("<strong>Industry sector</strong>"),
      pal = sector_palette, values = map_points$Sector,
      position = "bottomleft", opacity = 0.85
    ) |>
    addLegend(
      pal = decile_pal, values = factor(1:10),
      title = "IMD decile<br>(1 = most deprived)",
      position = "topright"
    ) |>
    
    addScaleBar(position = "bottomright") |>
    addLayersControl(
      overlayGroups = c("Point sources", "IMD decile", "Population density"),
      options = layersControlOptions(collapsed = FALSE)
    ) |>
    addEasyButton(easyButton(
      icon = "fa-list",
      title = "Show/hide legend",
      onClick = htmlwidgets::JS(" function(btn,map){ $('.legend').toggle(); } ")
    ))
}

# Example, once small_area_boundaries / pop_grid are loaded:
humberside_combined_map <- industrial_deprivation_map(
  centre_lat = 53.6545, centre_lng = -0.2310,
  boundaries = small_area_boundaries, pop_grid = pop_grid, imd_lookup = imd_uk
)


humberside_combined_map
```

```{r}
industrial_deprivation_map(
  centre_lat = 54.610, centre_lng = -1.168,
  boundaries = small_area_boundaries, pop_grid = pop_grid, imd_lookup = imd_uk
)

```

```{r}

industrial_deprivation_map(
  centre_lat = 51.710, centre_lng = -5.037,
  boundaries = small_area_boundaries, pop_grid = pop_grid, imd_lookup = imd_uk
)


```


```{r}


industrial_deprivation_map(
  centre_lat = 53.2738, centre_lng = -2.84,
  boundaries = small_area_boundaries, pop_grid = pop_grid, imd_lookup = imd_uk
)


```



```{r}



industrial_deprivation_map(
  centre_lat =50.84, centre_lng = -1.362,
  boundaries = small_area_boundaries, pop_grid = pop_grid, imd_lookup = imd_uk
)

```

```{r}

industrial_deprivation_map(
  centre_lat =56.020, centre_lng = -3.694,
  boundaries = small_area_boundaries, pop_grid = pop_grid, imd_lookup = imd_uk
)

```